Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers

### 2024-03-30 2_SSL_NE/article/article_main.tex
various updates

% number all lines as pre journal submission guidelines
\usepackage{lineno}
\linenumbers

\newcommand{\hlcolor}[2][yellow]{ {\sethlcolor{#1} \hl{#2}} }

@@

\begin{abstract}
\hl{
Since their introduction to the market in 1996, white light-emitting diodes (LEDs) have greatly improved in performance, efficiency, and manufacturing cost. Understanding the extent and mechanisms of the rapid progress in white LED technology can provide valuable insights for accelerating innovation in other demand-side clean energy technologies critical for reducing global carbon emissions. Here we show, through cost and performance modeling based on data from literature review, patent analysis, and expert interviews, that the efficiency of top-performing warm white GaN-based LED packages increased from 5.8\% in 2003 to 38.8\% in 2020. Over the same period, the manufacturing cost of low-to-mid-power LED packages decreased by 95.5\% from \$1.1 to \$0.05 (in 2020 USD). We find that technology spillovers from other sectors accounted for at least 8.5\% of efficiency improvements and nearly all consumer experience enhancements, playing an important role in widespread LED adoption in lighting.
}
\end{abstract}

@@

A rapid reduction of global carbon dioxide emissions is urgently required in order to mitigate the effects of climate change \cite{Forster2019}. While the Paris Agreement of 2015 requires signatory states to develop national emissions reduction strategies, these are not legally binding \cite{bodansky2016legal}. However, some countries have introduced more ambitious legislation, including the European Union, which has set a target of net zero emissions by 2050. It aims to meet this goal through an ambitious set of policies known as the European Green Deal \cite{eu2020green}.

Achieving such ambitious and critically important targets will require both the deployment of new clean energy technologies \cite{iea2020cleanenergy}, and the acceleration of innovation in existing supply-side \cite{sinn2012green} and demand-side energy technologies \cite{rgeVorsatz2009}. To ensure rapid adoption of these technologies, significant reductions in their costs and improvements in their performance and consumer experience are needed. This requires understanding how these cost reductions and performance improvements can be achieved \cite{Stephan2021,Ziegler2021}.

There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{WRIGHT_1936, Arrow_1962}, and economies of scope \cite{johansson2012global, national2016power, iea2020perspectives}. The role of these factors at different stages of the innovation life cycle, from research and technology development to demonstration, market formation and diffusion \cite{grubler2012policies}, is an area of active research  \cite{kavlak2018evaluating, Ziegler2021}. Innovation is also driven by forces of supply and demand \cite{Mowery1979}. Various \textit{"technology-push"} drivers reduce the costs of innovating, while \textit{"market-pull"} drivers increase the pay-offs from investing in innovation \cite{anadon2009policy}.

The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies, Anadon2016}. Among these drivers, the role of external knowledge and technology spillovers in research and development of energy technologies remains insufficiently studied \cite{Stephan2021}. While the exact definition of spillovers in the literature depends on the context \cite{Nemet2012, kolesnikov2021spillovers}, we specifically follow the approach of two previous studies \cite{Stephan2021, kolesnikov2021spillovers} by considering technology spillovers as the application of external knowledge in a technology where knowledge is considered external if it has been developed for application in other technologies, sectors, or scientific disciplines. There is emerging evidence that understanding spillovers and the knowledge network beyond a particular technology may be an important factor in understanding \cite{Pichler2020, iea2020cleanenergy} and shaping \cite{Clark2016, Stephan2021} the future evolution of technologies.

Among demand-side technologies, the provision of lighting is a particularly important area for climate change mitigation efforts, as it currently   accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}. It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996 \cite{Nakamura2013}, lighting technology has experienced dramatic efficiency improvements.  As shown in  \cref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}. For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W. Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices. \cref{fgr:cost_lamp_small} shows how LED retail prices have fallen by two orders of magnitude, at an annual price per flux decline of 27.3\% during the 2008-2020 period, in line with previous estimates \cite{Gerke2020}.

\begin{figure}[H]
 \centering
 \includegraphics[width=\textwidth]{figures/history_efficacy.pdf}
 \caption{\textbf{Historical progress in the luminous efficacy ($\eta$) of the most widely-used lighting technologies in human history.} Data points indicate best performers by year of market introduction. Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Note 2 for details). \hlcolor[green]{Dashed lines are guides to the eye, based on a 3rd-order polynomial fit to the data trend.} The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al. \cite{Murphy2012}. The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year. For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines. Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot. Abbreviations: HID - High-Intensity Discharge; CFL - Compact Fluorescent Lamp; Hal. - Halogen, Incd. - Incandescent. Source: own synthesis of published data based on a visual approach proposed by Azevedo et al. \cite{azevedo2009transition}. See Supplementary Note 16 for the full list of sources and references.}
 \label{fgr:history_efficacy}
 \vspace{-7mm}
\end{figure}

\begin{figure}[H]
\centering
  \includegraphics[height=5cm]{figures/cost_lamp_small.pdf}
  \caption{\textbf{Historical development of retail sales prices per luminous flux for LED-based luminaires in the period 2008-2020.} Luminaires considered include light bulbs, spotlights and recessed lights. Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S. Department of Energy (DOE) \cite{national2013assessment}. Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs, with the latter assumed constant based on the average in the covered time period. Source: own synthesis of data on sales prices collected from various consumer watchdog databases and publications. See Supplementary Note 16 for the full list of sources and references.}
  \label{fgr:cost_lamp_small}
\end{figure}

@@

Despite this impressive history, the sources of technological progress in white LEDs have not received as much attention from researchers as innovation in supply-side energy technologies, such as solar photovoltaics \cite{kavlak2018evaluating, nemet2019solar}, wind energy \cite{qiu2012price, jennings2020policy}, or lithium-ion batteries \cite{Ziegler2021, Stephan2021}. To the best of our knowledge, no study has comprehensively discussed the sources or extent of historical progress across a range of metrics related to LED cost and performance since the introduction of the first commercial white LED products (see Supplementary Note 1 for a brief review of previous literature on this topic). This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012, Creutzig2018}. Understanding the extent to which individual innovations and technology spillovers contributed to improvements in white LED technology, how their effect compares to other sources of improvements such as economies of scale, and how these innovations and spillovers occurred (i.e., by what mechanisms and actors) will provide valuable lessons for innovation in other demand-side energy technologies and clean energy innovation in general.

To address these questions, we use three sets of metrics to quantify historical progress in LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, specifically the colour rendering index (CRI) and colour temperature (CT); and 3) LED chip manufacturing cost. The rationale for the choice of these metrics, along with their detailed descriptions and definitions, are discussed in Supplementary Note 2. We find that  total LED device efficiency increased from 5.8\% to 38.8\% between 2003 and 2020, of which at least 8.5\% can be attributed specifically to technology spillovers. Our interviews have also revealed the important role of incremental manufacturing process improvements and learning-by-doing \cite{WRIGHT_1936, Arrow_1962} in this progress in LED efficiency. However, further research is needed to quantify the contribution of learning-by-doing to this progress. On the other hand, we find that technology spillovers contributed to nearly all consumer experience improvements in LED-based lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications. Finally, our LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid-power classic-chip-architecture GaN-based phosphor-converted warm-white LEDs from 1.11\$ to 0.05\$ (in 2020 USD) between 2003 and 2020. In contrast with LED performance improvements, where progress was driven mostly by innovations in LED technology, the dramatic decline in LED manufacturing cost has been a product of higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size. Overall we find that technology spillovers are most prevalent in those device loss channels that are well understood at the physical level. Improvements in device loss channels that are governed by complex quantum effects at the atomic level, and can only be described heuristically, have come from incremental manufacturing process improvements and learning-by-doing. Directing research efforts to better understand fundamental physical or chemical processes related to novel clean energy technologies is therefore likely to increase the possibility of technology spillovers.

@@

\hlcolor[green]{Using the mixed-methods approach described in Methods, we have collected data on the historical progress in individual sub-efficiencies of warm white phosphor-converted GaN-based LED devices, along with white LED innovations that have driven this progress. The results of our data collection efforts are presented in Supplementary Figure 7 and discussed in Supplementary Notes section 4.}Using this information, we calculate the overall lamp efficiency $\eta_L$ for the best performing mid/high-power phosphor-converted warm white LED devices in four years: 2003, 2010, 2016 and 2020. The waterfall diagrams of electric power input losses in \cref{fgr:waterfall} show how improvements in individual sub-efficiencies led to improvements in the overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to 12.7\% in 2010, 32.5\% in 2016 and finally to 38.8\% in 2020. As is evident from the figure, no single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations \cite{tsao2010solid}. We note, however, that the loss channels with a fixed physical limit on efficiency, e.g., Stokes loss that determines the light conversion efficiency by phosphors, became relatively more dominant in 2016 and 2020 compared to 2003 and 2010.  This is a direct result of large efficiency improvements in upstream sub-efficiencies.

@@

In \cref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of the corresponding loss channels. We find that there has been consistent progress across all device sub-efficiencies in the recorded period. Specifically, between 2003 and 2020, forward voltage efficiency increased from 70\% to 99.5\%, internal quantum efficiency from 55\% to 90\%, electrical droop from 65\% to 90\%, light extraction efficiency from 60\% to 90\%, spectral efficiency from 74\% to 83\%, conversion efficiency (red) from 11\% to 45\%, conversion efficiency (green) from 19\% to 61\%. Notably, some sub-efficiencies for the most recent devices considered in our study are now within $\sim10\%$ of their respective physical limits. The exception is spectral efficiency which, at $\sim17\%$ below the physical limit, shows larger potential for further improvements, which is important, given that efforts to improve white LED performance across different efficiency loss channels through R\&D continue \cite{cho2017white, Weisbuch2020}.

@@

\begin{table}[H]
    \footnotesize
    \centering
    \caption{\textbf{LED innovations related to improvements in consumer experience metrics.}}
    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|X|l|}
    \hline
        \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} & \textbf{Ref.} \\ \hline
        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 3.2.2.1 \\ \hline
        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs & SI 3.2.2.1 \\ \hline
        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 3.2.2.2 \\ \hline
        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 3.2.2.4 \\ \hline
        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 3.2.2.3 \\ \hline
        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 3.2.2.2 \\ \hline
        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 3.2.2.2 \\ \hline
    \end{NiceTabularX}
    \caption*{Note: \textit{Year} column represents the earliest reported application of innovation in white LEDs. These differ from the years shown in \cref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for LED products that relied on those innovations. Detailed descriptions of the history of each innovation and relevant spectral data are provided in Supplementary Note 4. Desig.=Designation, Ref.=Reference.}
    \label{tab:phosphors}
\end{table}

@@

Surprisingly, R\&D efforts to improve LED technology and associated technology spillovers are not found to have a substantial impact on LED manufacturing cost reductions over time. There are several explanations for this finding. First, the functional unit of our study is the LED chip itself, not the luminaire containing multiple chips. As LED efficiency and overall brightness have increased, a smaller number of chips can be used in luminaires to obtain the same level of the overall light flux. However, since we are investigating improvements at the level of single chips, we do not capture the cost effects downstream of chip manufacturing. We also previously showed that overall brightness is no longer a target metric in chip development \cite{weinold2021compound}. Second, we find that increases in the wafer size and higher overall manufacturing yield, which are jointly responsible for a 95.5\% manufacturing cost reduction per LED chip, were both enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing, rather than particular innovations in LED technology associated with R\&D efforts, as confirmed by interviews with experts in LED manufacturing.

@@

Among enabling factors that we identified for these spillovers, we highlight public mission-driven R\&D funding; industry-academia partnerships; firm experience in multiple industries; conferences that brought together researchers from academia and industry; cultural and language proximity; freedom of search in academia; and university alumni networks.
Finally, we find that, on average, it took 26 years from the initial invention or discovery taking place in a field of science and technology different from white LED lighting to the first time it got applied in (‘spilled over to’) the white LED lighting domain, with this time varying from 5 to almost 70 years. In contrast, it took much less time – just 6 years on average, varying from just a few months to 19 years – to develop a commercial application for the spillover knowledge in the LED lighting market.

@@

In this study, we use a multi-method approach to synthesize evidence from multiple sources and, for the first time, reconstruct a comprehensive picture of the rapid technological progress in white LED technology throughout its history since the introduction to the market in 1996 and across an ensemble of metrics related to device performance, cost, and consumer experience. Improvements in these metrics are traced to sources in innovation, technology spillovers, and economies of scale in the manufacturing process. Improvements in device efficiency are shown in \cref{fgr:waterfall} and \cref{fgr:breakthroughs_efficiency}. Improvements in consumer experience are shown in \cref{fgr:consumer_experience}. Improvements in manufacturing cost are shown in \cref{fgr:costmodel}.

>>>>

% number all lines as pre journal submission guidelines
\usepackage{lineno}
%\linenumbers

@@

\begin{abstract}

Since their introduction to the market in 1996, white light-emitting diodes (LEDs) have greatly improved in performance, efficiency, and manufacturing cost. Understanding the extent and mechanisms of the rapid progress in white LED technology can provide valuable insights for accelerating innovation in other demand-side clean energy technologies critical for reducing global carbon emissions. Here we show, through cost and performance modeling based on data from literature review, patent analysis, and expert interviews, that the efficiency of top-performing warm white GaN-based LED packages increased from 5.8\% in 2003 to 38.8\% in 2020. Over the same period, the manufacturing cost of low-to-mid-power LED packages decreased by 95.5\% from \$1.1 to \$0.05 (in 2020 USD). We find that technology spillovers from other sectors accounted for at least 8.5\% of efficiency improvements and nearly all consumer experience enhancements, playing an important role in widespread LED adoption in lighting.

\end{abstract}

@@

A rapid reduction of global carbon dioxide emissions is urgently required in order to mitigate the effects of climate change \cite{Forster2019}. While the 2015 Paris Agreement requires signatory states to develop national emissions reduction strategies, these are not legally binding \cite{bodansky2016legal}. However, some countries have introduced more ambitious legislation, including the European Union, which has set a target of net zero emissions by 2050. It aims to meet this goal through an ambitious set of policies known as the European Green Deal \cite{eu2020green}.

Achieving such ambitious and critically important targets requires both developing new clean energy technologies \cite{iea2020cleanenergy} and accelerating the deployment of existing supply-side \cite{sinn2012green} and demand-side energy technologies \cite{rgeVorsatz2009}. To ensure rapid adoption of these technologies, significant reductions in their costs and improvements in their performance are needed. This requires understanding how these cost reductions and performance improvements can be achieved \cite{nemet2019solar,Stephan2021,kavlak2018evaluating,Ziegler2021}.

Improvements in technology cost and performance, which in this study we generally call “innovations”, can occur via different mechanisms, including (but not limited to) learning by doing\cite{WRIGHT_1936, Arrow_1962}, targeted research and development (R\&D) efforts, or economies of scale\cite{johansson2012global, iea2020perspectives}. The role of these mechanisms at different stages of technology life cycle, from research and development to demonstration, market formation and diffusion\cite{grubler2012policies}, is an area of active research \cite{kavlak2018evaluating,Ziegler2021}. Different mechanisms can result in different kinds of innovations, characterized by what changes in the technology (e.g., its architecture, materials, components, or manufacturing process) and the scale of resulting improvements (e.g., radical or incremental) \cite{Acemoglu_2022}. For example, R\&D efforts (sometimes also called “learning-by-researching”) \cite{Castelnuovo_2005} more often than other mechanisms can result in substantial or even radical improvements in technology performance. On the other hand, learning-by-doing is typically associated with continuous incremental improvements in the technology resulting from accumulated experience in technology demonstration, manufacturing, and use. Both mechanisms, as well as economies of scale, can also lead to radical (i.e., discontinuous) or incremental reductions in technology costs.

Importantly, different mechanisms of innovation in a particular technology play out within a broader innovation system comprising multiple actors, networks, and institutions\cite{grubler2012policies, Anadon2016}. As a result, these mechanisms are affected by various drivers and barriers, including forces of supply and demand\cite{Mowery1979}, public policies, and regulation\cite{anadon2009policy}.

Among these drivers, the role of external knowledge and technology spillovers in research and development of energy technologies remains insufficiently studied \cite{Stephan2021}. While the exact definition of spillovers in the literature depends on the context \cite{Nemet2012, kolesnikov2021spillovers}, we specifically follow the approach of two previous studies \cite{Stephan2021, kolesnikov2021spillovers}. We define technology spillovers as the application of external knowledge in a technology where knowledge is considered external if it has been developed for application in other technologies, sectors, or scientific disciplines. There is emerging evidence that understanding spillovers and the knowledge network beyond a particular technology may be an important factor in understanding \cite{Pichler2020, iea2020cleanenergy} and shaping \cite{Clark2016, Stephan2021} the future evolution of technologies. For example, spillovers can provide critical external knowledge inputs to the R\&D process \cite{kolesnikov2021spillovers}, enable the re-use of experience from different sectors in learning-by-doing\cite{iea2020cleanenergy}, and lead to cost reductions, e.g., via repurposing the manufacturing equipment from other sectors\cite{Stephan2021}.

The aim of this paper is to identify and, where possible, quantify the contribution of different mechanisms and technology spillovers to the historical progress in a specific technology – white light-emitting diode (LED) lighting. Among demand-side technologies, general illumination is a particularly important area for climate change mitigation efforts, as it currently accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}. It is also an area of rapid recent technological change: since the introduction of first commercial white LEDs in 1996 \cite{Nakamura2013}, lighting technology has experienced dramatic efficiency improvements. As shown in \cref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}. For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W. Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices. \cref{fgr:cost_lamp_small} shows how LED retail prices have fallen by two orders of magnitude, at an annual price per flux decline of 27.3\% during the 2008-2020 period, in line with previous estimates \cite{Gerke2020}.

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{figures/history_efficacy.pdf}
 \caption{\textbf{Historical progress in the luminous efficacy ($\eta$) of the most widely-used lighting technologies in human history.} Data points indicate best performers by year of market introduction. Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Note 2 for details). Dashed lines are guides to the eye, based on a 3rd-order polynomial fit to the data trend. The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al. \cite{Murphy2012}. The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year. For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines. Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot. Abbreviations: HID - High-Intensity Discharge; CFL - Compact Fluorescent Lamp; Hal. - Halogen, Incd. - Incandescent. Source: own synthesis of published data based on a visual approach proposed by Azevedo et al. \cite{azevedo2009transition}. See Supplementary Note 16 for the full list of sources and references.}
 \label{fgr:history_efficacy}
\end{figure}

\begin{figure}[h!]
\centering
  \includegraphics[height=5cm]{figures/cost_lamp_small.pdf}
  \caption{\textbf{Historical development of retail sales prices per luminous flux for LED-based luminaires in the period 2008-2020.} Luminaires considered include light bulbs, spotlights and recessed lights. Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S. Department of Energy (DOE) \cite{national2013assessment}. Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs, with the latter assumed constant based on the average in the covered time period. Note that projections are consistently at the lower end of the retail price data. This is because retail price data includes more expensive luminaire types, while projections are for traditional light bulbs only. Projections also do not include value added tax. Source: own synthesis of data on sales prices collected from various consumer watchdog databases and publications. See Supplementary Note 16 for the full list of sources and references.}
  \label{fgr:cost_lamp_small}
\end{figure}

@@

Despite this impressive history, the mechanisms of technological progress in white LEDs have not received as much attention from researchers as in supply-side energy technologies, such as solar photovoltaics \cite{kavlak2018evaluating, nemet2019solar}, wind energy \cite{qiu2012price, jennings2020policy}, or lithium-ion batteries \cite{Ziegler2021, Stephan2021}. To the best of our knowledge, no study has comprehensively discussed the mechanisms or extent of historical progress across a range of metrics related to LED cost and performance since the introduction of the first commercial white LED products (see Supplementary Notes section 1 for a brief review of previous literature on this topic). This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012, Creutzig2018}. Understanding the extent to which individual innovations and technology spillovers contributed to overall improvements in white LED technology cost and performance, by which mechanism these innovations and spillovers occurred, and what factors affected this process will provide valuable lessons for innovation in other demand-side energy technologies and clean energy innovation in general.

To address these questions, we first identify three sets of metrics to quantify historical progress in LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, specifically the colour rendering index (CRI) and colour temperature (CT); and 3) LED chip manufacturing cost. The rationale for the choice of these metrics, along with their detailed descriptions and definitions, are discussed in Supplementary Notes section 2. Next, using data from literature review, patent analysis, and expert interviews, we trace the historical improvements in these metrics in warm white GaN-based phosphor-converted LEDs—currently the most widely adopted variant of white LED technology which we thus choose as the focal technological area for our analysis—from their introduction to the lighting market in 2003 to 2020, the year with the most recent data available at the time of data collection. We find that total LED device efficiency of such LEDs increased from 5.8\% to 38.8\% between 2003 and 2020. A majority of this increase can be attributed to a series of innovations in white LED technology resulting from R\&D efforts, with at least 8.5\% of the cumulative increase attributed specifically to technological innovations driven by technology spillovers. Our interviews also revealed a lesser but still important role of incremental manufacturing process improvements most likely resulting from learning-by-doing. However, substantial further research is needed to quantify the relative contribution of learning-by-doing vis-a-vis technological innovations to this progress. On the other hand, we find that innovations driven by technology spillovers contributed to nearly all consumer experience improvements in white LED lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications. Finally, our bottom-up LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid-power classic-chip-architecture GaN-based phosphor-converted warm-white LEDs from 1.11\$ to 0.05\$ (in 2020 USD) between 2003 and 2020. In contrast with LED performance improvements, where progress has been driven mostly by R\&D efforts, the dramatic decline in LED manufacturing cost has been a product of learning-by-doing resulting in higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size. We provide the details of our findings for the progress in white LED efficiency, cost, and consumer experience, as well as the role of technology spillovers in this progress, in the remaining sections below.

@@

Using the mixed-methods approach described in Methods, we have collected data on the historical progress in individual sub-efficiencies of warm white phosphor-converted GaN-based LED devices, along with white LED innovations that have driven this progress. The results of our data collection efforts are presented in Supplementary Figure 7 and discussed in Supplementary Notes section 4. Using this information, we calculate the overall lamp efficiency $\eta_L$ for the best performing mid/high-power phosphor-converted warm white LED devices in four years: 2003, 2010, 2016 and 2020. The waterfall diagrams of electric power input losses in \cref{fgr:waterfall} show how improvements in individual sub-efficiencies led to improvements in the overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to 12.7\% in 2010, 32.5\% in 2016 and finally to 38.8\% in 2020. As is evident from the figure, no single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations \cite{tsao2010solid}. We note, however, that the loss channels with a fixed physical limit on efficiency, e.g., Stokes loss that determines the light conversion efficiency by phosphors, became relatively more dominant in 2016 and 2020 compared to 2003 and 2010.  This is a direct result of large efficiency improvements in upstream sub-efficiencies.

@@

In \cref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of the corresponding loss channels. The estimates shown here for the contributions of different channels do not show uncertainty ranges because we lack reliable information about the distribution of input parameters, which would require us to arbitrarily assign range values for inputs, and the fact that we focus on a best case scenario. Thus, our point estimates should be interpreted as the expected values in a best case scenario. We find that there has been consistent progress across all device sub-efficiencies in the recorded period. Specifically, between 2003 and 2020, forward voltage efficiency increased from 70\% to 99.5\%, internal quantum efficiency from 55\% to 90\%, electrical droop from 65\% to 90\%, light extraction efficiency from 60\% to 90\%, spectral efficiency from 74\% to 83\%, conversion efficiency (red) from 11\% to 45\%, conversion efficiency (green) from 19\% to 61\%. Notably, some sub-efficiencies for the most recent devices considered in our study are now within $\sim10\%$ of their respective physical limits. The exception is spectral efficiency which, at $\sim17\%$ below the physical limit, shows larger potential for further improvements, which is important, given that efforts to improve white LED performance across different efficiency loss channels through R\&D continue \cite{cho2017white, Weisbuch2020}.

@@

\begin{table}[H]
    \footnotesize
    \centering
    \caption{\textbf{LED innovations related to improvements in consumer experience metrics.}}
    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|X|}
    \hline
        \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} \\ \hline
        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs \\ \hline
        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs \\ \hline
        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor \\ \hline
        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion \\ \hline
        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor \\ \hline
        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor \\ \hline
        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor\\ \hline
    \end{NiceTabularX}
    \caption*{Note: \textit{Year} column represents the earliest reported application of innovation in white LEDs. These differ from the years shown in \cref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for LED products that relied on those innovations. Detailed descriptions of the history of each innovation and relevant spectral data are provided in Supplementary Note 4. Desig.=Designation, Ref.=Reference.}
    \label{tab:phosphors}
\end{table}

@@

Surprisingly, R\&D efforts to improve LED technology and associated technology spillovers are not found to have a substantial impact on LED manufacturing cost reductions over time. There are several explanations for this finding. First, the functional unit of our study is the LED chip itself, not the luminaire containing multiple chips. As LED efficiency and overall brightness have increased, a smaller number of chips can be used in luminaires to obtain the same level of the overall light flux. However, since we are investigating improvements at the level of single chips, we do not capture the cost effects downstream of chip manufacturing. We also previously showed that overall brightness is no longer a target metric in chip development \cite{weinold2021compound}. Second, we find that an increase in wafer size and higher overall yield, taken together, are jointly responsible for a 95.5\% reduction of manufacturing cost per LED chip between 2003 and 2020. Both were enabled by advances in manufacturing equipment (e.g., in epitaxy) enabling the economies of scale in manufacturing, as well as incremental manufacturing process improvements from learning-by-doing, rather than LED technology innovations stemming from R\&D, which was confirmed by our interviews with experts in LED manufacturing.

@@

Among the enabling factors that we identified for these spillovers, we highlight public mission-driven R\&D funding; industry-academia partnerships; firm experience in multiple industries; conferences that brought together researchers from academia and industry; cultural and language proximity; freedom of search in academia; and university alumni networks.
Finally, we find that, on average, it took 26 years from the initial invention or discovery taking place in a field of science and technology different from white LED lighting to the first time it got applied in (‘spilled over to’) the white LED lighting domain, with this time varying from 5 to almost 70 years. In contrast, it took much less time – just 6 years on average, varying from just a few months to 19 years – to develop a commercial application for the spillover knowledge in the LED lighting market.

@@

In this study, we use a multi-method approach to synthesize evidence from multiple sources and, for the first time, reconstruct a detailed picture of the rapid technological progress in white LED technology throughout its history since the introduction to the market in 1996 and across an ensemble of metrics related to device performance (\cref{fgr:waterfall} and \cref{fgr:breakthroughs_efficiency}), manufacturing cost (\cref{fgr:costmodel}), and consumer experience (\cref{fgr:consumer_experience}). Improvements in these metrics are traced (with contributions quantified, where possible) to specific white LED technology innovations resulting from R\&D activities, technology spillovers, learning-by-doing, and economies of scale in the manufacturing process.

Specifically, we find that progress in LED efficiency and consumer experience has been predominantly driven by innovations resulting from R\&D activities, but the relative contribution of technology spillovers to these innovations has been very different: at least 8.5\% for efficiency improvements and virtually all 100\% for consumer experience metrics. In addition, we observe (but so far have not been able to quantify) a likely small but meaningful role of learning-by-doing in efficiency improvements. Interestingly, we find that technology spillovers are most prevalent in those device loss channels that are well understood at the physical level (e.g., light extraction and spectral efficiencies, see Fig. 4 and Table 2). We find that improvements in device loss channels that are governed by complex quantum effects at the atomic level (e.g., droop) and at present can only be described heuristically, have, on the other hand, come mostly from incremental manufacturing process improvements resulting from learning-by-doing with little evidence for spillovers.

In contrast with efficiency and consumer experience improvements, we find that dramatic manufacturing cost reductions in white LEDs have been mostly driven by learning-by-doing allowing higher yields across manufacturing steps, and economies of scale resulting from increases in the sapphire wafer size. We estimate a combined contribution of these two factors to cost reductions for warm white GaN-based phosphor-converted LEDs to be at least 95.5\%, leaving only a minor residual of 4.5\% for all other mechanisms of innovation, including R\&D.
